---
title: "Does more precision help DTW-SOM? (float → double → long double → quad → 50 digits → MPFR)"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>")
```

The core is templated on the number type, so the *same* code runs at every precision.
Grid geometry, schedules and neighbourhood weights stay in `double`; only the DTW
arithmetic and the prototypes use the chosen type.

| `precision`  | type                                  | mantissa | notes                                    |
|--------------|---------------------------------------|----------|------------------------------------------|
| `float`      | IEEE binary32                          | 24 bits  | ~7 digits                                |
| `double`     | IEEE binary64 (default)                | 53 bits  | ~16 digits                               |
| `longdouble` | x87 80-bit (gcc on x86)                | 64 bits  | ~19 digits; = double on arm64 macOS      |
| `quad`       | Boost `cpp_bin_float<113>` (binary128) | 113 bits | ~34 digits, software                     |
| `bin50`      | Boost `cpp_bin_float<50>`              | 168 bits | 50 digits, software                      |
| `mpfr100`    | Boost `mpfr_float_backend<100>`        | 333 bits | needs `-DDTWSOM_WITH_MPFR -lmpfr -lgmp`  |

Boost headers come from the `BH` R package, so nothing has to be installed at
system level. MPFR is optional (Rmpfr users already have the libraries; on
Windows they ship with Rtools).

```{r build}
library(dtwsom)
# MPFR build: install with  Sys.setenv(PKG_CXXFLAGS = "-DDTWSOM_WITH_MPFR", PKG_LIBS = "-lmpfr -lgmp")
lanes_src <- system.file("examples", "dtw_lanes.cpp", package = "dtwsom")   # for the vectorisation report below
str(dtwsom_build_info_cpp())
precs <- dtwsom_precisions_cpp(); precs
```

## 1. One DTW distance, all precisions

Three kinds of input, each designed to stress the arithmetic differently:

* **scaled**: z-normalised, length 300 - the normal case;
* **offset**: the same signal sitting on a DC offset of 10⁶ (raw ADC counts, unfiltered
  pressure or temperature sensors look like this) - cancellation in `(a - b)²`;
* **long**: length 5 000 - many additions in the cumulative cost.

The 50-digit result is the reference; the table shows the relative error of every
other precision against it, and the time per distance.

```{r ladder}
set.seed(1)
mk <- function(n, offset = 0) matrix(offset + sin(seq(0, 6 * pi, length.out = n)) + rnorm(n, sd = 0.3))
cases <- list(scaled = list(mk(300), mk(280)),
              offset = list(mk(300, 1e6), mk(280, 1e6)),
              long   = list(mk(5000), mk(4900)))

ladder <- function(a, b) {
  do.call(rbind, lapply(precs, function(p) {
    r <- dtw_distance_cpp(a, b, -1L, Inf, p)
    # time the plain distance-matrix path (dtw_distance_cpp also computes a 50-digit reference)
    t <- system.time(for (i in 1:5) dtw_cross_cpp(list(a), list(b), -1L, Inf, 1L, p))[["elapsed"]] / 5
    data.frame(precision = p, bits = r$mantissa_bits, value = substr(r$digits, 1, 36),
               rel_error_vs_50digits = r$rel_error_vs_bin50, ms = 1000 * t)
  }))
}
for (nm in names(cases)) { cat("\n==", nm, "==\n"); print(ladder(cases[[nm]][[1]], cases[[nm]][[2]]), digits = 3, row.names = FALSE) }
```

Read the `rel_error` column: `double` sits at 10⁻¹⁶ in every case, `float` at 10⁻⁷ on
scaled data and 10⁻³ on the offset data (it has only ~7 digits and 10⁶ eats 6 of them).
`longdouble` and `quad` add digits that no downstream decision can see. Note the cost
column: the software types are 20-100× slower.

## 2. Does precision change any *decision*?

A SOM only uses distances to (a) choose the BMU and (b) form averages. So the question
is not "are the digits different" but "does any BMU flip". Map the synthetic motif set
(and its 10⁶-offset twin) onto one fixed codebook at every precision and compare BMUs
against the 50-digit reference.

```{r motifs}
make_motifs <- function(n = 400, seed = 1, offset = 0) {
  set.seed(seed)
  shapes <- c("sine", "triangle", "square", "ramp")
  lab <- sample(shapes, n, replace = TRUE)
  x <- lapply(lab, function(s) {
    L <- sample(40:90, 1); u <- seq(0, 1, length.out = L); a <- runif(1, 0.7, 1.3)
    y <- switch(s, sine = sin(2 * pi * u), triangle = 1 - 4 * abs(u - 0.5),
                square = ifelse(u < 0.5, 1, -1), ramp = 2 * u - 1)
    offset + a * y + rnorm(L, sd = 0.15)
  })
  list(x = x, label = lab)
}
d  <- make_motifs()
d6 <- make_motifs(offset = 1e6)
set.seed(42); codes <- d$x[sample(400, 25)]; codes6 <- lapply(codes, `+`, 1e6)

bmu_agreement <- function(x, codes) {
  ref <- dtwsom_map_cpp(codes, x, 10L, Inf, -1L, 0L, "bin50")
  gap <- apply(ref$distmat, 1, function(r) { s <- sort(r)[1:2]; (s[2] - s[1]) / s[1] })  # relative margin of the BMU
  do.call(rbind, lapply(precs, function(p) {
    t <- system.time(m <- dtwsom_map_cpp(codes, x, 10L, Inf, -1L, 0L, p))[["elapsed"]]
    flips <- which(m$bmu != ref$bmu)
    data.frame(precision = p, bmu_flips = length(flips),
               smallest_margin_flipped = if (length(flips)) signif(min(gap[flips]), 2) else NA,
               max_rel_dist_err = { ok <- ref$distmat > 0; signif(max(abs(m$distmat[ok] - ref$distmat[ok]) / ref$distmat[ok]), 2) },
               seconds = round(t, 2))
  }))
}
cat("scaled data (margin between best and 2nd-best unit: median", signif(median(apply(dtwsom_map_cpp(codes, d$x, 10L, Inf, -1L, 0L)$distmat, 1, function(r) { s <- sort(r)[1:2]; (s[2]-s[1])/s[1] })), 2), ")\n")
print(bmu_agreement(d$x, codes), row.names = FALSE)
cat("\nsame data on a 1e6 offset\n")
print(bmu_agreement(d6$x, codes6), row.names = FALSE)
```

A BMU can only flip when the margin between the best and the second-best unit is smaller
than the arithmetic error. With `double` the error is 10⁻¹⁶ and the margins are 10⁻²…10⁻¹,
so the two never meet. `float` on offset data is the one case that breaks.

## 3. Train the same map at two precisions

Same initial codebook, same seed, same schedule - `double` versus `quad` (and `float`).
The batch algorithm is deterministic, so any difference is purely arithmetic.

```{r train, results = "hold"}
fit_at <- function(p, x = d$x) dtw_som(x, 3, 3, epochs = 8, mode = "batch", window = 10,
                                       codes = codes[1:9], precision = p, seed = 7, verbose = FALSE)
fits <- lapply(c(float = "float", double = "double", quad = "quad"), fit_at)
cmp <- function(a, b) c(
  same_bmu   = mean(a$unit.classif == b$unit.classif),
  max_rel_code_diff = max(mapply(function(u, v) max(abs(u - v)), a$codes, b$codes)) / max(abs(unlist(b$codes))),
  qe_rel_diff = (a$qe - b$qe) / b$qe, te_diff = a$te - b$te)
rbind(float_vs_quad  = cmp(fits$float,  fits$quad),
      double_vs_quad = cmp(fits$double, fits$quad))
sapply(fits, function(f) c(qe = f$qe, te = f$te, seconds = f$elapsed))
```

```{r train-offset}
fits6 <- lapply(c(float = "float", double = "double", quad = "quad"),
                function(p) dtw_som(d6$x, 3, 3, epochs = 8, window = 10, codes = codes6[1:9],
                                    precision = p, seed = 7, verbose = FALSE))
rbind(float_vs_quad  = cmp(fits6$float,  fits6$quad),
      double_vs_quad = cmp(fits6$double, fits6$quad))
```

`double` reproduces the `quad` map to ~10⁻¹⁴ (relative) in the prototypes and to the last
unit in the assignments; `float` is off by 10⁻³ on scaled data and destroys the map on the
offset data. If you ever *do* see double and quad
disagree on real data, the cause will be a genuine tie (two prototypes at equal distance),
which no precision fixes - it is a property of the data, and worth flagging in the output.

## 4. What the extra precision costs

```{r cost}
cost <- do.call(rbind, lapply(intersect(precs, c("float", "double", "longdouble", "quad", "bin50", "mpfr100")), function(p) {
  f <- fit_at(p); data.frame(precision = p, seconds = round(f$elapsed, 2), qe = signif(f$qe, 6))
}))
cost$relative <- round(cost$seconds / cost$seconds[cost$precision == "double"], 1)
print(cost, row.names = FALSE)
```

## 5. AVX-512 on a Ryzen 9950X - what it is and how to switch it on

**It is the same instruction set as on Intel.** AMD did not invent anything separate:
Zen 4 (7000 series) added AVX-512 with a double-pumped 256-bit datapath, Zen 5 (e.g. the
9950X) executes full 512-bit vectors. Ironically the mainstream Intel desktop parts
(12th-14th gen) have AVX-512 fused off, so an AMD desktop is currently the easy way to
get it. There is nothing AMD-specific to learn; there are three things to know:

1. **The compiler must be told the target.** `-march=native` (or `-march=znver5`, GCC ≥ 14;
   `znver4` on GCC 13 / Rtools 4.4) turns AVX-512 on. Check what your build got:

    ```{r flags}
    unlist(dtwsom_build_info_cpp()[c("avx512f", "avx2", "fma", "compiler")])
    ```

   Put `CXX17FLAGS = -O3 -march=native` in `~/.R/Makevars` (Windows: `Documents/.R/Makevars`).
   For code that must run on other machines use GCC's `__attribute__((target_clones("avx512f","avx2","default")))`
   on Linux, or a manual `__builtin_cpu_supports("avx512f")` switch on Windows (function
   multi-versioning needs ELF ifunc, which Windows lacks).

2. **GCC prefers 256-bit vectors by default** on most targets (a leftover from Intel's
   AVX-512 clock throttling). Add `-mprefer-vector-width=512` to get 64-byte vectors, and
   *measure* - on Zen 5 it usually helps, on older Intel it often does not. The compiler tells
   you what it did with `-fopt-info-vec-optimized`:

    ```{r vecreport}
    cxx <- strsplit(system2(file.path(R.home("bin"), "R"), c("CMD", "config", "CXX17"), stdout = TRUE), " ")[[1]]
    rep <- system2(cxx[1], c(cxx[-1], "-O3 -march=native -mprefer-vector-width=512 -fopenmp -fopt-info-vec-optimized",
                             "-c", lanes_src, "-o", tempfile(fileext = ".o"), paste0("-I", R.home("include")),
                             paste0("-I", system.file("include", package = "Rcpp"))), stdout = TRUE, stderr = TRUE)
    unique(grep("dtw_lanes.cpp.*vectorized using", rep, value = TRUE))   # the `#pragma omp simd` lane loop
    ```

3. **The DTW recursion does not vectorise as written.** `cur[j]` needs `cur[j-1]`, a
   loop-carried dependency, so `-march` alone changes nothing in the core. The fix is to lay
   the data out so that consecutive memory holds *different prototypes at the same time step*
   and run them in lockstep - then the loop over prototypes has no dependency and
   `#pragma omp simd` vectorises it. `src/dtw_lanes.cpp` is that kernel, in 30 lines and
   without intrinsics:

    ```{r lanes}
    set.seed(1); n <- 1000; L <- 64      # dtw_lanes_cpp / dtw_scalar_cpp are exported by the package
    P <- matrix(rnorm(L * n), L, n)      # 64 prototypes of length 1000, one per row
    q <- rnorm(n)
    stopifnot(all.equal(dtw_lanes_cpp(P, q), dtw_scalar_cpp(P, q)))
    bench <- function(f, reps = 5) system.time(for (i in seq_len(reps)) f())[["elapsed"]] / reps
    t <- c(scalar = bench(function() dtw_scalar_cpp(P, q)),
           lanes_double = bench(function() dtw_lanes_cpp(P, q)),
           lanes_float = bench(function() dtw_lanes_cpp(P, q, "float")))
    round(rbind(seconds = t, speedup = t["scalar"] / t), 3)
    ```

   The lane kernel needs same-length prototypes (bucket the codebook by length, or pad) and
   applies no band yet; this is the Phase-2 item in the plan. Try it on a Zen 5 machine with
   `-mprefer-vector-width=512` versus without - the report above says whether you got
   64- or 32-byte vectors, the table says whether it mattered.

## 6. Conclusions

* **`double` is not the bottleneck.** Its error (10⁻¹⁶) is nine orders of magnitude below the
  smallest margin that decides a BMU; quad/50-digit/MPFR reproduce the double map exactly at
  30-100× the cost. Keep them as a *diagnostic* (`precision = "quad"` on a small subset when
  a result looks suspicious), not as a production setting.
* **`float` is the interesting direction, not MPFR:** it halves memory traffic and doubles SIMD
  width, and it is exact enough on scaled data. Two rules make it safe: z-normalise (or at
  least remove the DC offset) before DTW, and accumulate the per-epoch means in double
  (already the case for the QE statistics).
* **AVX-512 pays only after the loop is restructured** (lanes across prototypes); with the
  serial recursion it is a no-op. That restructuring gives ~10-20× on its own; the 256 → 512
  bit step on top of it is machine-dependent and must be measured.
